---
title: "Behavioral Pattern Clustering"
output: html_notebook
date: "2026-04-20"
---

Load data

```{r}
#load cleaned csv
df <- read.csv("Rape_Data_cleaned.csv", header=TRUE)
head(df)

```

Process
```{r}
#Preprocessing coordinates
df$X_COORD_CD <- as.numeric(gsub(",", "", df$X_COORD_CD))
df$Y_COORD_CD <- as.numeric(gsub(",", "", df$Y_COORD_CD))


#Coordinate data frame with just coordinates
cluster_vars <- c("X_COORD_CD", "Y_COORD_CD")
coords <- df[,cluster_vars]
```

Feature Engineering 
```{r}
#create binary attributes

#df$hour <- as.numeric(df$hour)
df$is_night <- ifelse(df$TIME_OF_DAY == "Night", 1,0)
df$is_inside <- ifelse(df$LOC_OF_OCCUR_DESC == "INSIDE", 1, 0)
df$is_completed <- ifelse(df$CRM_ATPT_CPTD_CD == "COMPLETED", 1, 0) 
df$is_weekend <- ifelse(df$DAY_OF_WEEK == "Saturday" | df$DAY_OF_WEEK == "Sunday", 1, 0)
#residential, public???


#treat precinct as a category 
#df$ADDR_PCT_CD <- as.factor(df$ADDR_PCT_CD)


```

```{r}
#library(dplyr)
#create separate df with time, location, and completion status per precinct
precinct_df <- aggregate(cbind(is_night, is_inside, is_completed) ~ ADDR_PCT_CD, data = df, FUN=mean, na.rm=TRUE)
head(precinct_df)
str(precinct_df)

#scale data so numbers are equal distance 
colnames(precinct_df) <- c("ADDR_PCT_CD", "is_night", "is_inside", "is_completed")
scaled_df <- scale(precinct_df[,-1])

```

Elbow Graph for Precinct Clusters

```{r}
set.seed(42)

wss <- sapply(2:15, function(k) {
  kmeans(scaled_df, centers = k, nstart = 25)$tot.withinss
})

plot(2:15, wss, type = "b", pch = 19,
     xlab = "Number of Clusters (k)",
     ylab = "Total Within-Cluster Sum of Squares (WSS)",
     main = "Elbow Plot for Optimal K (NYC Rape Incident Behavioral Patterns)")

# Choose optimal K at bend -> k=4
```


```{r}
# create clusters
k_km  <- 4
kmeans_result <- kmeans(scaled_df, centers = k_km, nstart = 25)
km_groups <- kmeans_result$cluster

#Add attribute to data
precinct_df$cluster <- kmeans_result$cluster
table(precinct_df$cluster) #quantity of each group
```
Understand Clusters
```{r}
#cluster information
aggregate(precinct_df[, c("is_night", "is_inside", "is_completed")], by= list(cluster=precinct_df$cluster), mean)
```
Cluster Summary

Cluster 1: strong nighttime and indoor pattern 
Cluster 2: mostly indoor, mixed timing (day/evening/night)
Cluster 3: no nighttime incidents, no inside classification, all completed cases *extremely rare case
Cluster 4: very strong indoor pattern, lower completion rate, mixed timing 

Most incidents are INSIDE (0.82-0.94) -> nighttime varies by cluster
HIGH overall completion rate (>0.92)

Visualize Clusters
```{r}
#pca visualization
library(ggplot2)

princ <- prcomp(scale(precinct_df[,c("is_night", "is_inside", "is_completed")]))
pca <- as.data.frame(princ$x[,1:2])
pca$cluster <- as.factor(precinct_df$cluster)

ggplot(pca, aes(x=PC1, y= PC2, color = cluster)) + 
  geom_point(size=3, alpha=0.8) + 
  labs(title = "K-means Clusters of NYC Incident Patterns (PCA Projection)",
       x= "Principal Component 1", y = "Principal Component 2")

```
```{r}
#inside frequency by cluster 
avg_inside <- aggregate(is_inside ~ cluster, data = precinct_df, mean)

ggplot(avg_inside, aes(x=factor(cluster), y=is_inside)) + geom_col(fill="navy") + labs(title = "Average Indoor Rate by Cluster", x="Cluster", y="Proportion Inside") 


```
```{r}
cluster_counts <- as.data.frame(table(precinct_df$cluster))
colnames(cluster_counts) <- c("cluster", "count")

ggplot(cluster_counts, aes(x=factor(cluster), y=count)) + geom_col(fill="darkred") + labs(title = "Number of Precincts per Cluster", x="Cluster", y="Count") 
```
```{r}
write.csv(df, "Rape_Data_Behavioral_Clusters.csv", row.names = FALSE)
```

Geographical Mapping
```{r}
#load data
library(sf)
# read shapefile
shapefile <- st_read("~/Downloads/tl_2025_us_county/tl_2025_us_county.shp") #from US CENSUS
#extract NY -> FIBS=36
ny_data <- shapefile[shapefile$STATEFP == "36",]
nrow(ny_data)

#extract NYC
  # Bronx = 36005
  # Brooklyn = 36047
  # Manhattan = 36061
  # Queens = 36081
  # Staten Island = 36085

nyc <- ny_data[ny_data$GEOID %in% c("36005","36047","36061","36081","36085"), ]

```
Convert Lat/Long Points
```{r}
#load NYC Rape data
rape_data <- read.csv("Rape_Data_clusters.csv", header=TRUE)
rape_points <- st_as_sf(rape_data, coords = c("Longitude","Latitude"), crs = 4326) #crs = 4326 - GPS Location
rape_points <- st_transform(rape_points, st_crs(nyc))
```

NYC Map Outline
```{r}
#change county name -> borough name
nyc$BOROUGH[nyc$GEOID == "36005"] <- "Bronx"
nyc$BOROUGH[nyc$GEOID == "36047"] <- "Brooklyn"
nyc$BOROUGH[nyc$GEOID == "36061"] <- "Manhattan"
nyc$BOROUGH[nyc$GEOID == "36081"] <- "Queens"
nyc$BOROUGH[nyc$GEOID == "36085"] <- "Staten Island"

map_outline <- ggplot(data=nyc) + geom_sf(fill = "lightgrey", color = "black") + geom_sf_label(aes(label = BOROUGH))

rape_points <- merge(rape_points,
                   precinct_df[, c("ADDR_PCT_CD", "cluster")],
                   by = "ADDR_PCT_CD",
                   all.x = TRUE)

ggplot() + geom_sf(data=nyc, fill="white", color="black") + geom_sf(data = rape_points, aes(color=factor(cluster)), alpha=0.5) +
  scale_color_manual(values = c("1" = "salmon2", "2" = "forestgreen", "3" = "turquoise3", "4" ="plum"))+
  labs(title="Georgaphical Location of Precinct Behavioral Profiles Clusters", color="Cluster")
```

